---
title: "Introduction to piping"
author: "Suzan Baert"
date: "25/01/2018"
output:
  xaringan::moon_reader:
    lib_dir: libs
    # nature:
    #   highlightStyle: github
    #   highlightLines: true
    #   countIncrementalSlides: false
---

## What is piping?

When you want to wrangle a large dataframe, you have a few options

1. Nest a lot of functions
  
  
  

``` {r eval=FALSE}
pizza <- take_out(put_in(dress_with(dress_with
        (dress_with(pie, sauce), oil), cheese), oven), oven)
```

.footnote[
*I found this nice example on someone's blog, but google does not find it back... * `r icon::fa("meh-o")`
]

---

## What is piping?

When you want to wrangle a large dataframe, you have a few options

1. Nest a lot of functions
1. Create a lot of temporary objects 
  
  

``` {r eval=FALSE}
pie_tmp <- dress_with(pie, sauce)
pie_tmp <- dress_with(pie_tmp, oil)
pie_tmp <- dress_with(pie_tmp, cheese)
pie_tmp <- put_in(pie_tmp, oven)
pizza <- take_out(pie_tmp, oven)

```

.footnote[
*I found this nice example on someone's blog, but google does not find it back... * `r icon::fa("meh-o")`
]
---

## What is piping?

When you want to wrangle a large dataframe, you have a few options

1. Nest a lot of functions
2. Create a lot of temporary variables
3. Use pipes  
  

``` {r eval=FALSE}
pizza <- pie %>% 
  dress_with(sauce) %>% 
  dress_with(oil) %>% 
  dress_with(break(cheese)) %>% 
  put_in(oven) %>%
  take_out(oven)

```

.footnote[
*I found this nice example on someone's blog, but google does not find it back... * `r icon::fa("meh-o")`
]
---

## What is piping?

When use the pipe-operator you build a chain of operations:

``` {r eval=FALSE}
dataframe %>% 
  function() %>% 
  function(args) %>% 
  function(args)

```


**Advantages of piping**
- structuring sequences of data operations left-to-right, making it more readable for humans
- avoiding nested function calls,
- minimizing the need for local variables 
- making it easy to add steps anywhere in the sequence of operations
- making it easy to stepwise explore someone elses code
---

